A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-06 podc_paper/boundedspace.tex
some editing of GC

To prevent RBTs from growing without bound, we would like to discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic.
We msut keep a block $B$ if any element enqueued by the operations in $E(B)$ has
not yet been dequeued, since a later dequeue may need to traverse the block to 
find its response.
Because of the FIFO nature of queues, this means that we can throw away 
blocks whose indices are smaller than the block containing the oldest enqueue whose
value is still in the queue.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than a given threshold from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
After every $p^2$ successful insertions into the RBT representing $\var{root}.\fld{blocks}$,
processes cooperate to perform \op{Split} operations on all nodes of the tournament tree  
to discard unneeded blocks.
We keep an array $\var{last}[1..p]$ where each process writes the index of the last block in
\var{root} that it dequeued an element from.
If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
of $p^2$, it does GC by first helping all pending dequeue operations by other processes,
then reading $\var{last}[1..p]$.

\here{need to complete the description of GC}

@@

\paragraph{Garbage Collection}
We did not handle garbage collection: \nf{Enqueue} operations  remain in the nodes even after their elements have been dequeued.
We can keep track of the \nf{block}s in the \nf{root} whose operations are all terminated, i.e., all enqueues have been dequeued, and the responses of all dequeues have been computed. We call these blocks \it{finished blocks}. If we help the operations of all processes to compute their responses, then we can say if block $B$ is finished, then all blocks before $B$ are also finished. Knowing the most recent finished block in a node, we can reclaim the memory taken by finished blocks. We cannot use arrays (or vectors) to throw the garbage blocks away. We need a data structure that supports \nf{tryAppend()}, \nf{read(i)}, \nf{write(i)} and \nf{split(i)} operations in $O(\log n)$ time, where \nf{split(i)} removes all the indices less than~\nf{i}. If each process tries to do the garbage collection once every $p^2$ operations on the queue, then the amortized complexity remains the same. We can use a concurrent implementation of a persistent red-black trees for this~\cite{DBLP:conf/afp/Okasaki96}. Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way.

\begin{figure}[hbt]  
  \center\includegraphics[width=3.5in]{pics/finishedBlocks.png}
  \caption[Blocks that can be safely garbage collected.]{Finished blocks are shown with red color and unfinished blocks are shown with green color. All the subblocks of a finished block are also finished.}
  \label{fig::finishedBlock}
\end{figure}

>>>>

To prevent RBTs from growing without bound, we would like to discard
blocks that are no longer needed.
Ensuring the size of the RBT is polynomial in $p$ and $q$ will 
also keep the running time of our operations polylogarithmic.
Blocks should be kept if they contain operations still in progress.
Moreover, a block containing an \op{Enqueue}($x$) must be kept until $x$ is dequeued.

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $root.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the tournament tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

\here{Have to talk about repeating CAS to do GC at each node up to p times}

Because of GC, an operation may not find a block $B$ that it requires when it searches a RBT.  If the operation
is a dequeue, it must have been helped before $B$ was discarded, so it can simply read its response from
its own leaf.
\here{What about enqueues not finding required block?}

@@
